---
title: "Airports by country"
author: "Nina De Grandis"
date: "2025-04-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## What countries have the most airports?

```{r}
head(airports_clean)

country_count <- airports_clean %>%
  count(country) %>%
  arrange(desc(n)) 

print(country_count)

top_25_countries <- country_count %>%
  arrange(desc(n)) %>%
  head(25)

top_25_countries
```

```{r}
ggplot(top_25_countries, aes(x = reorder(country, n), y = n)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +
  labs(title = "Top 25 Countries by Count", 
       x = "Country", 
       y = "Count") +
  theme_minimal()

```

### Comments

This bar chart shows the top 25 countries by number of airports, and how many airports they have. I chose to only shows the top 25 because any more countries made the graph very hard to read. There are 235 countries in the data set, and everything below the top 25 has only 45 or less airports. The United States takes a staggering lead, with 1251 airports. The next closest is Canada, with 380. Only the top 10 countries have 100 or more airports. These top 10 countries are among the most populated countries in the world, which would make sense why they have the most airports.

## Where are the top countries by airports located?

```{r}
airport_count_by_country <- airports_clean %>%
  count(country) %>%
  arrange(desc(n)) %>%
  head(25)

top_25_airports <- airports_clean %>%
  filter(country %in% airport_count_by_country$country)

library(maps)

ggplot() +
  borders("world", colour = "gray", fill = "lightgray") +  
  geom_point(data = top_25_airports, aes(x = longitude, y = latitude), 
             alpha = 0.5, color = "blue", size = 0.05) +
  labs(title = "Top 25 Countries by Number of Airports", 
       x = "Longitude", 
       y = "Latitude") +
  theme_minimal()

```

### Comments

This map shows where the airports for the top 25 countries by number of airports are located. I notice several trends when looking at this map. First, many of the airports are along the coast. This makes sense as these may be more populated areas, especially when accounting for tourism and economic activity. When talking about population, this may be why some areas are blank and without an airport for miles. Australia for example, has most of their population centered in big cities and along the coasts. An airport in the middle of Australia isn't very necessary. This is similar to Russia, where there is a seeming lack of airports for the size of the country. Russia's climate and population distribution accounts for this. Another noticeable trend is that Africa is completely left out of the top 25, except for one country, being South Africa. Most every other region of the world has a country with many airports, making travel very accessible.
