---
title: "Global Air Traffic Concentration Analysis by Airport Concentration"
author: "Lucas Martins Sorge"
date: "2025-04-27"
output: html_document
---

## Introduction

This report investigates how global air traffic is distributed across airports using the OpenFlights dataset.\
We analyze the concentration of flights among airports by visualizing the Lorenz Curve, calculating the Gini coefficient, and identifying top-contributing airports.

------------------------------------------------------------------------

## Data Preparation

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ineq)

# Load cleaned flight traffic dataset
flight_traffic <- read_csv("../cleaned_data/flight_traffic.csv")
```

------------------------------------------------------------------------

## Airport Traffic Concentration Analysis

## Lorenz Curve of Airport Traffic

We first examine the distribution of total flight traffic among airports

```{r}
# Arrange airports by descending total flight traffic
# Calculate cumulative sums and percentages for Lorenz curve plotting
flight_traffic_sorted <- flight_traffic %>%
  arrange(desc(flight_traffic)) %>%
  mutate(
    cumulative_flights = cumsum(flight_traffic),
    total_flights = sum(flight_traffic),
    cumulative_percent = cumulative_flights / total_flights,
    airport_percent = row_number() / n()
  )

head(flight_traffic_sorted)

# Plot the Lorenz Curve
ggplot(flight_traffic_sorted, aes(x = airport_percent, y = cumulative_percent)) +
  geom_line(color = "black", linewidth = 1.2) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  labs(
    title = "Lorenz Curve of Global Air Traffic Concentration",
    x = "Cumulative % of Airports",
    y = "Cumulative % of Flights"
  )

ggsave(
  filename   = "../figures/lorenz_airport.png",
  plot       = last_plot(),
  width      = 6,
  height     = 4,
  units      = "in",
  dpi        = 300
)

```

### Interpretation

The Lorenz Curve is bowed significantly below the line of perfect equality, indicating a high concentration of air traffic among a small proportion of airports.

------------------------------------------------------------------------

## Gini Coefficient Calculation

We compute the Gini coefficient to numerically measure the inequality in flight distribution across airports.

```{r}
# Calculate Gini Coefficient to quantify inequality of flight distribution
gini_coefficient <- ineq(flight_traffic_sorted$flight_traffic, type = "Gini")
gini_coefficient
```

### Interpretation

The Gini coefficient is approximately **0.78**, confirming that global air traffic is highly concentrated among a few airports.

------------------------------------------------------------------------

## Flight Concentration by Airport Percentile

We analyze what proportion of total flights are handled by the top 1%, 5%, and 10% of airports.

```{r}
# Concentration at different thresholds
top_1_percent_threshold <- flight_traffic_sorted %>%
  filter(airport_percent <= 0.01) %>%
  summarise(percent_flights = max(cumulative_percent)) %>%
  pull(percent_flights)

top_5_percent_threshold <- flight_traffic_sorted %>%
  filter(airport_percent <= 0.05) %>%
  summarise(percent_flights = max(cumulative_percent)) %>%
  pull(percent_flights)

top_10_percent_threshold <- flight_traffic_sorted %>%
  filter(airport_percent <= 0.10) %>%
  summarise(percent_flights = max(cumulative_percent)) %>%
  pull(percent_flights)

cat("Top 1% of airports handle about", round(top_1_percent_threshold*100, 1), "% of flights.\n")
cat("Top 5% of airports handle about", round(top_5_percent_threshold*100, 1), "% of flights.\n")
cat("Top 10% of airports handle about", round(top_10_percent_threshold*100, 1), "% of flights.\n")
```

### Interpretation

-   The top **1%** of airports handle about **20%** of all flights.

-   The top **5%** of airports handle about **53%** of all flights.

-   The top **10%** of airports handle about **70%** of all flights.

These findings demonstrate the extreme concentration of air traffic at a small number of major hubs.

------------------------------------------------------------------------

## Top 10 Airports by Total Flights

Finally, we identify the top airports contributing the most to global traffic.

```{r}
# Top 10 airports by traffic
top_airports_table <- flight_traffic_sorted %>%
  slice_max(flight_traffic, n = 10) %>%
  select(iata, flight_traffic, cumulative_percent)

top_airports_table
```

### Interpretation

The busiest airports, including **Atlanta (ATL)**, **Chicago O'Hare (ORD)**, and **Beijing Capital (PEK)**, dominate the global air traffic network, accounting for a substantial proportion of total flights.

------------------------------------------------------------------------

## Conclusion

Our analysis shows that global air traffic is highly concentrated.

\
The Lorenz Curve and a high Gini coefficient (\~0.78) confirm that a small percentage of airports handle a disproportionately large share of flights.\

This concentration reflects the critical role of a limited number of international hubs in global aviation connectivity.
